---
title: "Exploring ANOM"
subtitle: ""
description: ""
authors: 
  - name: "Dan Swart, CPA (ret)"
  - name: "Claude Sonnet 4.5"
date: today
date-format: long
# bibliography: manual-refs.bib
format:
  html:
    resources:
      - reference-backlinks.js
    include-after-body:    
      - text: |
          # <script type="text/javascript" src="reference-backlinks.js"></script>
    default: true         
    code-copy: true
    code-link: true        # This adds individual buttons
    code-fold: true        # Hide code by default, show on click
    code-summary: "Show the code"
    code-overflow: wrap
    code-block-bg: "#FAEBD7"
    code-block-border-left: "#31BAE9"
    embed-resources: true
    include-in-header: header.html
    css:
      - swart.css
      - tachyons.min.css
      - r-colors.css
    fontsize: 18pt
    lightbox: true
    page-layout: full
    fig-width: 12
    fig-height: 10
    fig-dpi: 300
    html-math-method: katex
    df-print: paged
    toc: true
    toc-float: true
    citeproc: true
    link-citations: true
    linestretch: 1.0
    
    
    
  typst:
    fig-width: 12
    fig-height: 10
    fig-dpi: 300
    margin:
      x: 1in
      y: 1in
    toc: true
    fontsize: 14pt
    mainfont: "Cabin"
    
  revealjs:
    slide-number: true
    transition: fade
    code-overflow: wrap
    center: true
    smaller: true
    scrollable: true
    chalkboard: true
    multiplex: false
    theme: solarized
    reference-location: margin
    logo: img/red-cross-640-435.png
    footer: "Footer text"
    code-block-height: 650px



  # docx:
  #   highlight-style: github
  #   fig_caption: true



editor: source

quarto:
  render:
    cache-refresh: true


# for .qmd filesd
execute:
  echo: false
  message: false
  warning: false
  eval: true
  fig-width: 12
  fig-height: 10


# for .rmd files
knitr:
  opts_chunk:
    echo: false
    error: false
    warning: false
    message: false
    cache: false

---


```{r}
#| label: setup
#| include: false
# R version 4.4.3 (2025-02-28)
# Platform: x86_64-apple-darwin20
# Running under: macOS Sequoia 15.6.1
# Rstudio 2025.9.2.418 (Cucumberleaf Sunflower)
# 

# Prevent scientific notation globally
options(scipen = 999)
options(device = "quartz")  # macOS native graphics device
options(tigris_use_cache = TRUE)

# install.packages(c("mapview", "survey", "srvyr", "arcgislayers"))

# census_api_key("95496766c51541ee6f402c1e1a8658581285b759", install = TRUE, overwrite = TRUE)


# load libraries

# library(ggplot2) # Create Elegant Data Visualizations Using the Grammar of Graphics
# library(tibble) # Simple Data Frames
# library(tidyr) # Tidy Messy Data
# library(readr) # Read Rectangular Text Data
# library(purrr) # Functional Programming Tools
# library(dplyr) # A Grammar of Data Manipulation
# library(stringr) # Simple, Consistent Wrappers for Common String Operations
# library(forcats) # Tools for Working with Categorical Variables (Factors)
# library(lubridate) # Make Dealing with Dates a Little Easier
# library(readxl) # Read Excel Files
# library(writexl) # Export Data Frames to Excel 'xlsx' Format 
# library(janitor) # Simple Tools for Examining and Cleaning Dirty Data 
# library(ggtext) # Improved Text Rendering Support for 'ggplot2' 
# library(paletteer) # Comprehensive Collection of Color Palettes 
# library(viridis) # Colorblind-Friendly Color Maps for R 
# library(RColorBrewer) # ColorBrewer Palettes 
# library(wesanderson) # A Wes Anderson Palette Generator 
# library(dutchmasters) # Color Palettes based on Famous Paintings 
# library(gghighlight) # Highlight Lines and Points in 'ggplot2' 
# library(monochromeR) # Easily Create, View and Use Monochrome Color Palettes 
# library(ggforce) # Accelerating 'ggplot2' 
# library(ggthemes) # Extra Themes, Scales and Geoms for 'ggplot2'
# library(gt) # Easily Create Presentation-Ready Display Tables
# library(gtExtras) # Extending 'gt' for Beautiful HTML Tables
# library(plotly) # Create Interactive Web Graphics via 'plotly.js'
# library(patchwork) # The Composer of Plots
# library(ppcor) # Partial and Semi-Partial (Part) Correlation
# library(ggplot2) # Create Elegant Data Visualizations Using the Grammar of Graphics
# library(corrplot) # Visualization of a Correlation Matrix
# library(ggcorrplot) # Visualization of a Correlation Matrix using 'ggplot2'
# library(car) # Companion to Applied Regression
# library(WRS2) # A Collection of Robust Statistical Methods
# library(boot) # Bootstrap Functions
# library(BayesFactor) # Computation of Bayes Factors for Common Designs
# library(pwr) # Basic Functions for Power Analysis
# library(qgraph) # Graph Plotting Methods, Psychometric Data Visualization and Graphical Model Estimation
# library(scales) # Scale Functions for Visualization
# library(here) # A Simpler Way to Find Your Files
# library(ggdag) # Analyze and Create Elegant Directed Acyclic Graphs      
# library(dagitty) # Graphical Analysis of Structural Causal Models    
# library(DiagrammeR) # Graph/Network Visualization 
# library(knitr) # A General-Purpose Package for Dynamic Report Generation in R     
# library(kableExtra) # Construct Complex Table with 'kable' and Pipe Syntax
# library(flextable) # Functions for Tabular Reporting  
# library(lavaan) # Latent Variable Analysis     
# library(ggpubr) # 'ggplot2' Based Publication Ready Plots
# library(rethinking) # Statistical Rethinking book package
# library(broom) # Convert Statistical Objects into Tidy Tibbles
# library(tidycensus) # Load US Census Boundary and Attribute Data as 'tidyverse' and 'sf'-Ready Data Frames


# Force dplyr's select to take precedence
select <- dplyr::select
filter <- dplyr::filter

# Set global theme for consistent plots
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 20) + 
                   ggplot2::theme(plot.title = ggplot2::element_text(face = "bold",
                                                                      size = 26),
                                  plot.subtitle = ggplot2::element_text(face = "bold",
                                                                         size = 24),
                                  axis.title.x = ggplot2::element_text(face = "bold",
                                                                        size = 22),
                                  axis.title.y = ggplot2::element_text(face = "bold",
                                                                        size = 22),
                                  axis.text.x = ggplot2::element_text(face = "bold",
                                                                       size = 22,
                                                                       angle = 45,
                                                                       hjust = 1),
                                  legend.position = "bottom",
                                  strip.text = ggplot2::element_text(face = "bold"),
                                  panel.spacing.x = grid::unit(1.5,
                                                                "cm"),
                                  panel.spacing.y = grid::unit(1.5,
                                                                "cm"),
                                  plot.margin = ggplot2::margin(20,
                                                                 20,
                                                                 20,
                                                                 20,
                                                                 "pt")))


# Set seed for reproducibility
set.seed(123)

```


# Analysis of Means (ANOM)

## Two Methods of Creating the ANOM Chart

* With the first method we're comparing each group's mean to the overall mean (rather than directly comparing groups)

* We calculate decision limits based on the overall variance

* Groups with means outside these limits are considered significantly different from the average

* The visual representation focuses on whether groups differ from the overall mean

This approach is particularly useful when you have more than two groups, as it provides a single comprehensive analysis rather than multiple pairwise comparisons.


The key differences in the two types of Analysis of Means approaches are:


For Analysis of Means (ANOM), the type of values (percentages, raw scores, or proportions) does matter to the analysis, but with different considerations than other statistical tests.

The key considerations for ANOM with different types of data are:

- ***Raw scores***: Standard ANOM works well with continuous data that follows approximately normal distributions. The decision limits are based on the assumption of normality and equal variances across groups.

- ***Percentages/Proportions***: When working with percentages or proportions (values between 0 and 1):

- ***The variance is not constant*** but depends on the mean (binomial variance = p(1-p)/n)

- Standard ANOM assumptions are violated when proportions are close to 0 or 1

- ***Non-normal data***: For highly skewed data (which can happen with percentages), transformations may be needed before applying ANOM

When any of these conditions exist a transformed version called "ANOM for Proportions" should be used, which applies an arcsine transformation


## Standard ANOM Chart


```{r}
#| echo: false
#| warning: false

# Given summary statistics
male_mean <- 78.5
male_sd <- 12.3
female_mean <- 80.2
female_sd <- 11.6
male_n <- 500 # Estimated sample size
female_n <- 500 # Estimated sample size

# Calculate overall (grand) mean
overall_n <- male_n + female_n
overall_mean <- (male_mean * male_n + female_mean * female_n) / overall_n

# Calculate pooled standard deviation
pooled_sd <- sqrt(((male_n - 1) * male_sd^2 + (female_n - 1) * female_sd^2) /
                  (male_n + female_n - 2))

# Calculate standard error for each group
male_se <- pooled_sd / sqrt(male_n)
female_se <- pooled_sd / sqrt(female_n)

# Calculate critical value (for 95% confidence)
# Using t-distribution with df = n1 + n2 - 2
alpha <- 0.05
critical_value <- stats::qt(1 - alpha / 2,
                            df = male_n + female_n - 2)

# Calculate decision limits
upper_decision_limit <- overall_mean +
  critical_value * pooled_sd * sqrt(1 / male_n + 1 / female_n)
lower_decision_limit <- overall_mean -
  critical_value * pooled_sd * sqrt(1 / male_n + 1 / female_n)

# Create data frame for ANOM plot
anom_df <- data.frame(group = c("Male",
                                 "Female"),
                      mean = c(male_mean,
                               female_mean),
                      n = c(male_n,
                            female_n))

# Calculate if means are significantly different from overall mean
anom_df$significant <- (anom_df$mean > upper_decision_limit) |
  (anom_df$mean < lower_decision_limit)

# Create ANOM plot
ggplot2::ggplot(data = anom_df,
                mapping = ggplot2::aes(x = group,
                                       y = mean,
                                       color = significant)) +
  ggplot2::geom_point(size = 4) +
  ggplot2::geom_hline(yintercept = overall_mean,
                      linetype = "solid") +
  ggplot2::geom_hline(yintercept = upper_decision_limit,
                      linetype = "dashed") +
  ggplot2::geom_hline(yintercept = lower_decision_limit,
                      linetype = "dashed") +
  ggplot2::annotate(geom = "text",
                    x = 1.5,
                    y = overall_mean + 1,
                    label = paste("Overall Mean =",
                                  round(overall_mean,
                                        2))) +
  ggplot2::annotate(geom = "text",
                    x = 1.5,
                    y = upper_decision_limit + 1,
                    label = "Upper Decision Limit") +
  ggplot2::annotate(geom = "text",
                    x = 1.5,
                    y = lower_decision_limit - 1,
                    label = "Lower Decision Limit") +
  ggplot2::scale_color_manual(values = c("black",
                                         "red")) +
  ggplot2::labs(title = "Analysis of Means (ANOM)",
                subtitle = "Comparing Male and Female Score Means to Overall Mean",
                y = "Mean Score",
                x = "Group") +
  ggplot2::theme(legend.position = "none")

# Function to calculate overlap coefficient from Cohen's d
calculate_overlap <- function(d) {
  # Overlap coefficient formula
  ovl <- 2 * stats::pnorm(-abs(d) / 2) * 100
  return(ovl)
}

```

<br>
<br>

```{r}
#| echo: false
#| warning: false

# Calculate additional statistics for interpretation
stats_df <- data.frame(statistic = c("Overall Mean",
                                      "Male Mean",
                                      "Female Mean",
                                      "Difference (F-M)",
                                      "Cohen's d",
                                      "Distribution Overlap",
                                      "Upper Decision Limit",
                                      "Lower Decision Limit",
                                      "Significant Difference?"),
                       value = c(round(overall_mean,
                                       2),
                                 male_mean,
                                 female_mean,
                                 round(female_mean - male_mean,
                                       2),
                                 round((female_mean - male_mean) / pooled_sd,
                                       2),
                                 round(calculate_overlap((female_mean - male_mean) / pooled_sd),
                                       1),
                                 round(upper_decision_limit,
                                       2),
                                 round(lower_decision_limit,
                                       2),
                                 ifelse(any(anom_df$significant),
                                        "Yes",
                                        "No")))

# Create flextable
stats_df |>
  flextable::flextable() |>
  flextable::set_header_labels(statistic = "Statistic",
                                value = "Value") |>
  flextable::theme_booktabs() |>
  flextable::autofit() |>
  flextable::align(align = "left",
                   part = "all") |>
  flextable::bold(part = "header")
```

<br>


## ANOM Chart with Arcsine Transformation

This approach accounts for the special statistical properties of proportions, ensuring that the ANOM results are valid regardless of whether you're working with raw scores or percentage/proportion data.


```{r}
#| echo: false
#| warning: false

# Example with proportion data (e.g., pass rates)
male_prop <- 0.785 # 78.5% pass rate
female_prop <- 0.802 # 80.2% pass rate
male_n <- 500 # Sample size
female_n <- 500 # Sample size

# Calculate overall proportion
overall_n <- male_n + female_n
overall_prop <- (male_prop * male_n + female_prop * female_n) / overall_n

# Apply arcsine transformation (stabilizes variance for proportions)
male_trans <- asin(sqrt(male_prop))
female_trans <- asin(sqrt(female_prop))
overall_trans <- asin(sqrt(overall_prop))

# Standard error for transformed proportions
se_trans <- sqrt(1 / (4 * mean(c(male_n,
                                  female_n))))

# Critical value (for 95% confidence)
alpha <- 0.05
critical_value <- stats::qnorm(1 - alpha / 2)

# Decision limits for transformed scale
upper_limit_trans <- overall_trans + critical_value * se_trans
lower_limit_trans <- overall_trans - critical_value * se_trans

# Convert limits back to proportion scale for interpretation
upper_limit_prop <- sin(upper_limit_trans)^2
lower_limit_prop <- sin(lower_limit_trans)^2

# Create data frame for ANOM plot
anom_df <- data.frame(group = c("Male",
                                 "Female"),
                      proportion = c(male_prop,
                                     female_prop),
                      transformed = c(male_trans,
                                      female_trans),
                      n = c(male_n,
                            female_n))

# Check significance
anom_df$significant <- (anom_df$transformed > upper_limit_trans) |
  (anom_df$transformed < lower_limit_trans)

# Calculate spacing for annotations
y_range <- upper_limit_prop - lower_limit_prop
annotation_offset <- y_range * 0.08

# Plot on original proportion scale
ggplot2::ggplot(data = anom_df,
                mapping = ggplot2::aes(x = group,
                                       y = proportion,
                                       color = significant)) +
  ggplot2::geom_point(size = 4) +
  ggplot2::geom_hline(yintercept = overall_prop,
                      linetype = "solid") +
  ggplot2::geom_hline(yintercept = upper_limit_prop,
                      linetype = "dashed") +
  ggplot2::geom_hline(yintercept = lower_limit_prop,
                      linetype = "dashed") +
  ggplot2::annotate(geom = "text",
                    x = 1.5,
                    y = overall_prop + annotation_offset,
                    label = paste("Overall Proportion =",
                                  round(overall_prop,
                                        3))) +
  ggplot2::annotate(geom = "text",
                    x = 1.5,
                    y = upper_limit_prop + annotation_offset,
                    label = "Upper Decision Limit") +
  ggplot2::annotate(geom = "text",
                    x = 1.5,
                    y = lower_limit_prop - annotation_offset,
                    label = "Lower Decision Limit") +
  ggplot2::scale_color_manual(values = c("black",
                                         "red")) +
  ggplot2::labs(title = "ANOM for Proportions using Arcsine Transformation",
                subtitle = "Comparing Male and Female Proportions to Overall Proportion",
                y = "Proportion",
                x = "Group") +
  ggplot2::theme(legend.position = "none")

```

<br>
<br>

```{r}
#| echo: false
#| warning: false

# Calculate Cohen's h for proportions (effect size)
cohens_h <- 2 * (asin(sqrt(female_prop)) - asin(sqrt(male_prop)))

# Calculate additional statistics for interpretation
stats_df <- data.frame(statistic = c("Overall Proportion",
                                      "Male Proportion",
                                      "Female Proportion",
                                      "Difference (F-M)",
                                      "Cohen's h",
                                      "Upper Decision Limit",
                                      "Lower Decision Limit",
                                      "Significant Difference?"),
                       value = c(round(overall_prop,
                                       4),
                                 round(male_prop,
                                       4),
                                 round(female_prop,
                                       4),
                                 round(female_prop - male_prop,
                                       4),
                                 round(cohens_h,
                                       4),
                                 round(upper_limit_prop,
                                       4),
                                 round(lower_limit_prop,
                                       4),
                                 ifelse(any(anom_df$significant),
                                        "Yes",
                                        "No")))

# Create flextable
stats_df |>
  flextable::flextable() |>
  flextable::set_header_labels(statistic = "Statistic",
                                value = "Value") |>
  flextable::theme_booktabs() |>
  flextable::autofit() |>
  flextable::align(align = "left",
                   part = "all") |>
  flextable::bold(part = "header")

```
<br>
<br>


# Understanding Statistical Power in ANOM

To understand when ANOM will detect significant differences between groups, we need to examine how effect size, sample size, and variability interact. This section explores three scenarios with progressively larger differences to illustrate the threshold at which ANOM identifies groups as significantly different from the overall mean.

**Why This Matters**: In practice, small differences between groups may not be statistically detectable, even if they exist. Understanding the minimum detectable difference helps researchers design studies with adequate power and helps decision-makers interpret results appropriately.

<br>

## Scenario 1: Small Difference (Cohen's d = 0.10)

In this scenario, we maintain the same sample sizes (n=500 per group) but reduce the difference between male and female means to create a very small effect.

**Parameters adjusted**: Mean values only (Female mean reduced from 80.2 to 78.6)

```{r}
#| echo: false
#| warning: false

# Scenario 1: Small difference
male_mean_s1 <- 78.5
male_sd_s1 <- 12.3
female_mean_s1 <- 78.6  # Very small difference
female_sd_s1 <- 11.6
male_n_s1 <- 500
female_n_s1 <- 500

# Calculate overall mean
overall_n_s1 <- male_n_s1 + female_n_s1
overall_mean_s1 <- (male_mean_s1 * male_n_s1 + female_mean_s1 * female_n_s1) / overall_n_s1

# Calculate pooled SD
pooled_sd_s1 <- sqrt(((male_n_s1 - 1) * male_sd_s1^2 + (female_n_s1 - 1) * female_sd_s1^2) /
                     (male_n_s1 + female_n_s1 - 2))

# Calculate critical value
alpha <- 0.05
critical_value_s1 <- stats::qt(1 - alpha / 2,
                               df = male_n_s1 + female_n_s1 - 2)

# Calculate decision limits
upper_limit_s1 <- overall_mean_s1 +
  critical_value_s1 * pooled_sd_s1 * sqrt(1 / male_n_s1 + 1 / female_n_s1)
lower_limit_s1 <- overall_mean_s1 -
  critical_value_s1 * pooled_sd_s1 * sqrt(1 / male_n_s1 + 1 / female_n_s1)

# Create data frame
anom_df_s1 <- data.frame(group = c("Male",
                                    "Female"),
                         mean = c(male_mean_s1,
                                  female_mean_s1),
                         n = c(male_n_s1,
                               female_n_s1))

anom_df_s1$significant <- (anom_df_s1$mean > upper_limit_s1) |
  (anom_df_s1$mean < lower_limit_s1)

# Plot
ggplot2::ggplot(data = anom_df_s1,
                mapping = ggplot2::aes(x = group,
                                       y = mean,
                                       color = significant)) +
  ggplot2::geom_point(size = 4) +
  ggplot2::geom_hline(yintercept = overall_mean_s1,
                      linetype = "solid") +
  ggplot2::geom_hline(yintercept = upper_limit_s1,
                      linetype = "dashed") +
  ggplot2::geom_hline(yintercept = lower_limit_s1,
                      linetype = "dashed") +
  ggplot2::annotate(geom = "text",
                    x = 1.5,
                    y = overall_mean_s1 + 0.5,
                    label = paste("Overall Mean =",
                                  round(overall_mean_s1,
                                        2))) +
  ggplot2::annotate(geom = "text",
                    x = 1.5,
                    y = upper_limit_s1 + 0.5,
                    label = "Upper Limit") +
  ggplot2::annotate(geom = "text",
                    x = 1.5,
                    y = lower_limit_s1 - 0.5,
                    label = "Lower Limit") +
  ggplot2::scale_color_manual(values = c("black",
                                         "red")) +
  ggplot2::labs(title = "Scenario 1: Small Effect (Cohen's d ≈ 0.01)",
                subtitle = "Difference of 0.1 points - Not Detectable",
                y = "Mean Score",
                x = "Group") +
  ggplot2::theme(legend.position = "none")

```

<br>
<br>

```{r}
#| echo: false
#| warning: false

# Calculate Cohen's d for scenario 1
cohens_d_s1 <- (female_mean_s1 - male_mean_s1) / pooled_sd_s1

scenario1_stats <- data.frame(statistic = c("Male Mean",
                                            "Female Mean",
                                            "Difference",
                                            "Pooled SD",
                                            "Cohen's d",
                                            "Upper Limit",
                                            "Lower Limit",
                                            "Significant?"),
                              value = c(male_mean_s1,
                                       female_mean_s1,
                                       round(female_mean_s1 - male_mean_s1,
                                             2),
                                       round(pooled_sd_s1,
                                             2),
                                       round(cohens_d_s1,
                                             3),
                                       round(upper_limit_s1,
                                             2),
                                       round(lower_limit_s1,
                                             2),
                                       ifelse(any(anom_df_s1$significant),
                                              "Yes",
                                              "No")))

scenario1_stats |>
  flextable::flextable() |>
  flextable::set_header_labels(statistic = "Parameter",
                                value = "Value") |>
  flextable::theme_booktabs() |>
  flextable::autofit() |>
  flextable::align(align = "left",
                   part = "all") |>
  flextable::bold(part = "header")

```

<br>

**Result**: With only a 0.1-point difference, both group means fall well within the decision limits. This tiny effect size (Cohen's d ≈ 0.01) is far too small to be detected with these sample sizes.

<br>
<br>

## Scenario 2: Moderate Difference (Cohen's d = 0.14)

We increase the difference between groups to create a moderate effect while keeping variability constant.

**Parameters adjusted**: Mean values only (Female mean set to 80.2)

```{r}
#| echo: false
#| warning: false

# Scenario 2: Moderate difference
male_mean_s2 <- 78.5
male_sd_s2 <- 12.3
female_mean_s2 <- 80.2  # Original moderate difference
female_sd_s2 <- 11.6
male_n_s2 <- 500
female_n_s2 <- 500

# Calculate overall mean
overall_n_s2 <- male_n_s2 + female_n_s2
overall_mean_s2 <- (male_mean_s2 * male_n_s2 + female_mean_s2 * female_n_s2) / overall_n_s2

# Calculate pooled SD
pooled_sd_s2 <- sqrt(((male_n_s2 - 1) * male_sd_s2^2 + (female_n_s2 - 1) * female_sd_s2^2) /
                     (male_n_s2 + female_n_s2 - 2))

# Calculate critical value
critical_value_s2 <- stats::qt(1 - alpha / 2,
                               df = male_n_s2 + female_n_s2 - 2)

# Calculate decision limits
upper_limit_s2 <- overall_mean_s2 +
  critical_value_s2 * pooled_sd_s2 * sqrt(1 / male_n_s2 + 1 / female_n_s2)
lower_limit_s2 <- overall_mean_s2 -
  critical_value_s2 * pooled_sd_s2 * sqrt(1 / male_n_s2 + 1 / female_n_s2)

# Create data frame
anom_df_s2 <- data.frame(group = c("Male",
                                    "Female"),
                         mean = c(male_mean_s2,
                                  female_mean_s2),
                         n = c(male_n_s2,
                               female_n_s2))

anom_df_s2$significant <- (anom_df_s2$mean > upper_limit_s2) |
  (anom_df_s2$mean < lower_limit_s2)

# Plot
ggplot2::ggplot(data = anom_df_s2,
                mapping = ggplot2::aes(x = group,
                                       y = mean,
                                       color = significant)) +
  ggplot2::geom_point(size = 4) +
  ggplot2::geom_hline(yintercept = overall_mean_s2,
                      linetype = "solid") +
  ggplot2::geom_hline(yintercept = upper_limit_s2,
                      linetype = "dashed") +
  ggplot2::geom_hline(yintercept = lower_limit_s2,
                      linetype = "dashed") +
  ggplot2::annotate(geom = "text",
                    x = 1.5,
                    y = overall_mean_s2 + 0.5,
                    label = paste("Overall Mean =",
                                  round(overall_mean_s2,
                                        2))) +
  ggplot2::annotate(geom = "text",
                    x = 1.5,
                    y = upper_limit_s2 + 0.5,
                    label = "Upper Limit") +
  ggplot2::annotate(geom = "text",
                    x = 1.5,
                    y = lower_limit_s2 - 0.5,
                    label = "Lower Limit") +
  ggplot2::scale_color_manual(values = c("black",
                                         "red")) +
  ggplot2::labs(title = "Scenario 2: Moderate Effect (Cohen's d ≈ 0.14)",
                subtitle = "Difference of 1.7 points - Still Not Detectable",
                y = "Mean Score",
                x = "Group") +
  ggplot2::theme(legend.position = "none")

```

<br>
<br>

```{r}
#| echo: false
#| warning: false

# Calculate Cohen's d for scenario 2
cohens_d_s2 <- (female_mean_s2 - male_mean_s2) / pooled_sd_s2

scenario2_stats <- data.frame(statistic = c("Male Mean",
                                            "Female Mean",
                                            "Difference",
                                            "Pooled SD",
                                            "Cohen's d",
                                            "Upper Limit",
                                            "Lower Limit",
                                            "Significant?"),
                              value = c(male_mean_s2,
                                       female_mean_s2,
                                       round(female_mean_s2 - male_mean_s2,
                                             2),
                                       round(pooled_sd_s2,
                                             2),
                                       round(cohens_d_s2,
                                             3),
                                       round(upper_limit_s2,
                                             2),
                                       round(lower_limit_s2,
                                             2),
                                       ifelse(any(anom_df_s2$significant),
                                              "Yes",
                                              "No")))

scenario2_stats |>
  flextable::flextable() |>
  flextable::set_header_labels(statistic = "Parameter",
                                value = "Value") |>
  flextable::theme_booktabs() |>
  flextable::autofit() |>
  flextable::align(align = "left",
                   part = "all") |>
  flextable::bold(part = "header")

```

<br>

**Result**: Even with a 1.7-point difference (Cohen's d ≈ 0.14), the means still fall within the decision limits. The high variability (SD ≈ 12) relative to the mean difference prevents detection of significance.

<br>
<br>

## Scenario 3: Large Difference - Detecting Significance

To achieve significance, we need either a much larger mean difference or reduced variability. Here we demonstrate both approaches.

**Parameters adjusted**: Mean values (Female mean increased to 83.5) AND standard deviations (both reduced by 20% to 9.84 and 9.28)

```{r}
#| echo: false
#| warning: false

# Scenario 3: Larger difference + reduced variability
male_mean_s3 <- 78.5
male_sd_s3 <- 9.84  # Reduced by 20%
female_mean_s3 <- 83.5  # Larger difference
female_sd_s3 <- 9.28  # Reduced by 20%
male_n_s3 <- 500
female_n_s3 <- 500

# Calculate overall mean
overall_n_s3 <- male_n_s3 + female_n_s3
overall_mean_s3 <- (male_mean_s3 * male_n_s3 + female_mean_s3 * female_n_s3) / overall_n_s3

# Calculate pooled SD
pooled_sd_s3 <- sqrt(((male_n_s3 - 1) * male_sd_s3^2 + (female_n_s3 - 1) * female_sd_s3^2) /
                     (male_n_s3 + female_n_s3 - 2))

# Calculate critical value
critical_value_s3 <- stats::qt(1 - alpha / 2,
                               df = male_n_s3 + female_n_s3 - 2)

# Calculate decision limits
upper_limit_s3 <- overall_mean_s3 +
  critical_value_s3 * pooled_sd_s3 * sqrt(1 / male_n_s3 + 1 / female_n_s3)
lower_limit_s3 <- overall_mean_s3 -
  critical_value_s3 * pooled_sd_s3 * sqrt(1 / male_n_s3 + 1 / female_n_s3)

# Create data frame
anom_df_s3 <- data.frame(group = c("Male",
                                    "Female"),
                         mean = c(male_mean_s3,
                                  female_mean_s3),
                         n = c(male_n_s3,
                               female_n_s3))

anom_df_s3$significant <- (anom_df_s3$mean > upper_limit_s3) |
  (anom_df_s3$mean < lower_limit_s3)

# Plot
ggplot2::ggplot(data = anom_df_s3,
                mapping = ggplot2::aes(x = group,
                                       y = mean,
                                       color = significant)) +
  ggplot2::geom_point(size = 4) +
  ggplot2::geom_hline(yintercept = overall_mean_s3,
                      linetype = "solid") +
  ggplot2::geom_hline(yintercept = upper_limit_s3,
                      linetype = "dashed") +
  ggplot2::geom_hline(yintercept = lower_limit_s3,
                      linetype = "dashed") +
  ggplot2::annotate(geom = "text",
                    x = 1.5,
                    y = overall_mean_s3 + 0.5,
                    label = paste("Overall Mean =",
                                  round(overall_mean_s3,
                                        2))) +
  ggplot2::annotate(geom = "text",
                    x = 1.5,
                    y = upper_limit_s3 + 0.5,
                    label = "Upper Limit") +
  ggplot2::annotate(geom = "text",
                    x = 1.5,
                    y = lower_limit_s3 - 0.5,
                    label = "Lower Limit") +
  ggplot2::scale_color_manual(values = c("black",
                                         "red")) +
  ggplot2::labs(title = "Scenario 3: Large Effect (Cohen's d ≈ 0.53)",
                subtitle = "Difference of 5 points with reduced variability - SIGNIFICANT",
                y = "Mean Score",
                x = "Group") +
  ggplot2::theme(legend.position = "none")

```

<br>
<br>

```{r}
#| echo: false
#| warning: false

# Calculate Cohen's d for scenario 3
cohens_d_s3 <- (female_mean_s3 - male_mean_s3) / pooled_sd_s3

scenario3_stats <- data.frame(statistic = c("Male Mean",
                                            "Female Mean",
                                            "Difference",
                                            "Pooled SD",
                                            "Cohen's d",
                                            "Upper Limit",
                                            "Lower Limit",
                                            "Significant?"),
                              value = c(male_mean_s3,
                                       female_mean_s3,
                                       round(female_mean_s3 - male_mean_s3,
                                             2),
                                       round(pooled_sd_s3,
                                             2),
                                       round(cohens_d_s3,
                                             3),
                                       round(upper_limit_s3,
                                             2),
                                       round(lower_limit_s3,
                                             2),
                                       ifelse(any(anom_df_s3$significant),
                                              "Yes",
                                              "No")))

scenario3_stats |>
  flextable::flextable() |>
  flextable::set_header_labels(statistic = "Parameter",
                                value = "Value") |>
  flextable::theme_booktabs() |>
  flextable::autofit() |>
  flextable::align(align = "left",
                   part = "all") |>
  flextable::bold(part = "header")

```

<br>

**Result**: SUCCESS! By combining a larger mean difference (5 points) with reduced variability (20% decrease in SD), we achieve a Cohen's d of approximately 0.53. The female mean now falls outside the upper decision limit, indicating a statistically significant difference from the overall mean.

<br>
<br>

## Scenario Comparison Summary

```{r}
#| echo: false
#| warning: false

comparison_df <- data.frame(scenario = c("Scenario 1: Small",
                                         "Scenario 2: Moderate",
                                         "Scenario 3: Large"),
                            mean_diff = c(round(female_mean_s1 - male_mean_s1,
                                               2),
                                         round(female_mean_s2 - male_mean_s2,
                                               2),
                                         round(female_mean_s3 - male_mean_s3,
                                               2)),
                            pooled_sd = c(round(pooled_sd_s1,
                                               2),
                                         round(pooled_sd_s2,
                                               2),
                                         round(pooled_sd_s3,
                                               2)),
                            cohens_d = c(round(cohens_d_s1,
                                              3),
                                        round(cohens_d_s2,
                                              3),
                                        round(cohens_d_s3,
                                              3)),
                            significant = c("No",
                                          "No",
                                          "Yes"))

comparison_df |>
  flextable::flextable() |>
  flextable::set_header_labels(scenario = "Scenario",
                                mean_diff = "Mean Difference",
                                pooled_sd = "Pooled SD",
                                cohens_d = "Cohen's d",
                                significant = "Significant?") |>
  flextable::theme_booktabs() |>
  flextable::autofit() |>
  flextable::align(align = "center",
                   part = "body") |>
  flextable::align(align = "left",
                   j = 1,
                   part = "body") |>
  flextable::bold(part = "header") |>
  flextable::bg(i = ~ significant == "Yes",
                bg = "#90EE90",
                part = "body")

```

<br>
<br>

**Key Insights**:

1. **Effect size matters more than raw difference**: A 1.7-point difference isn't significant with SD ≈ 12, but a 5-point difference becomes significant when SD is reduced to ≈ 9.6

2. **The signal-to-noise ratio is crucial**: Statistical significance depends on the ratio of mean difference to variability (pooled SD), not just the absolute difference

3. **Practical vs. statistical significance**: Even the "non-significant" 1.7-point difference might be practically meaningful in some contexts, while a statistically significant 5-point difference might not matter in others

4. **Design implications**: To detect smaller effects, researchers can either increase sample sizes or work to reduce measurement variability

<br>
<br>

# When Standard and Arcsine Methods Diverge for Proportions

The arcsine transformation becomes increasingly important when working with extreme proportions (near 0 or 1) or unbalanced sample sizes. Let's examine three cases to see when the methods diverge.

## Case 1: Moderate Proportions (Standard Method Works Fine)

When proportions are in the middle range (0.3 to 0.7), both methods produce similar results.

```{r}
#| echo: false
#| warning: false

# Case 1: Moderate proportions
p1_case1 <- 0.45
p2_case1 <- 0.50
n1_case1 <- 500
n2_case1 <- 500

# Standard method (using normal approximation)
overall_p_case1 <- (p1_case1 * n1_case1 + p2_case1 * n2_case1) / (n1_case1 + n2_case1)
se_standard_case1 <- sqrt(overall_p_case1 * (1 - overall_p_case1) * (1/n1_case1 + 1/n2_case1))
z_crit <- stats::qnorm(1 - 0.05/2)
upper_standard_case1 <- overall_p_case1 + z_crit * se_standard_case1
lower_standard_case1 <- overall_p_case1 - z_crit * se_standard_case1

# Arcsine method
trans1_case1 <- asin(sqrt(p1_case1))
trans2_case1 <- asin(sqrt(p2_case1))
overall_trans_case1 <- asin(sqrt(overall_p_case1))
se_arcsin_case1 <- sqrt(1 / (4 * mean(c(n1_case1,
                                         n2_case1))))
upper_trans_case1 <- overall_trans_case1 + z_crit * se_arcsin_case1
lower_trans_case1 <- overall_trans_case1 - z_crit * se_arcsin_case1
upper_arcsin_case1 <- sin(upper_trans_case1)^2
lower_arcsin_case1 <- sin(lower_trans_case1)^2

# Create comparison data
method_compare_case1 <- data.frame(method = c("Standard",
                                               "Standard",
                                               "Arcsine",
                                               "Arcsine"),
                                   limit_type = c("Upper",
                                                  "Lower",
                                                  "Upper",
                                                  "Lower"),
                                   value = c(upper_standard_case1,
                                            lower_standard_case1,
                                            upper_arcsin_case1,
                                            lower_arcsin_case1))

method_compare_case1 |>
  flextable::flextable() |>
  flextable::set_header_labels(method = "Method",
                                limit_type = "Limit",
                                value = "Value") |>
  flextable::colformat_double(j = "value",
                               digits = 4) |>
  flextable::theme_booktabs() |>
  flextable::autofit() |>
  flextable::align(align = "left",
                   part = "all") |>
  flextable::bold(part = "header") |>
  flextable::add_header_row(values = c("",
                                       "Case 1: Moderate Proportions (p ≈ 0.45-0.50)"),
                             colwidths = c(1, 2))

```

<br>

**Observation**: The decision limits are nearly identical between methods (difference < 0.001). Both methods are reliable in this range.

<br>
<br>

## Case 2: Extreme Proportions (Methods Begin to Diverge)

When proportions approach 0 or 1, the variance becomes non-constant, and the methods begin to show differences.

```{r}
#| echo: false
#| warning: false

# Case 2: Extreme proportions
p1_case2 <- 0.05
p2_case2 <- 0.08
n1_case2 <- 500
n2_case2 <- 500

# Standard method
overall_p_case2 <- (p1_case2 * n1_case2 + p2_case2 * n2_case2) / (n1_case2 + n2_case2)
se_standard_case2 <- sqrt(overall_p_case2 * (1 - overall_p_case2) * (1/n1_case2 + 1/n2_case2))
upper_standard_case2 <- overall_p_case2 + z_crit * se_standard_case2
lower_standard_case2 <- overall_p_case2 - z_crit * se_standard_case2

# Arcsine method
trans1_case2 <- asin(sqrt(p1_case2))
trans2_case2 <- asin(sqrt(p2_case2))
overall_trans_case2 <- asin(sqrt(overall_p_case2))
se_arcsin_case2 <- sqrt(1 / (4 * mean(c(n1_case2,
                                         n2_case2))))
upper_trans_case2 <- overall_trans_case2 + z_crit * se_arcsin_case2
lower_trans_case2 <- overall_trans_case2 - z_crit * se_arcsin_case2
upper_arcsin_case2 <- sin(upper_trans_case2)^2
lower_arcsin_case2 <- sin(lower_trans_case2)^2

# Create comparison data
method_compare_case2 <- data.frame(method = c("Standard",
                                               "Standard",
                                               "Arcsine",
                                               "Arcsine"),
                                   limit_type = c("Upper",
                                                  "Lower",
                                                  "Upper",
                                                  "Lower"),
                                   value = c(upper_standard_case2,
                                            lower_standard_case2,
                                            upper_arcsin_case2,
                                            lower_arcsin_case2),
                                   difference = c(NA,
                                                 NA,
                                                 upper_arcsin_case2 - upper_standard_case2,
                                                 lower_arcsin_case2 - lower_standard_case2))

method_compare_case2 |>
  flextable::flextable() |>
  flextable::set_header_labels(method = "Method",
                                limit_type = "Limit",
                                value = "Value",
                                difference = "Diff from Standard") |>
  flextable::colformat_double(j = c("value", "difference"),
                               digits = 4) |>
  flextable::theme_booktabs() |>
  flextable::autofit() |>
  flextable::align(align = "left",
                   part = "all") |>
  flextable::bold(part = "header") |>
  flextable::add_header_row(values = c("",
                                       "",
                                       "Case 2: Extreme Proportions (p ≈ 0.05-0.08)"),
                             colwidths = c(1, 1, 2))

```

<br>

**Observation**: With extreme proportions near 0, the methods show noticeable differences (up to 0.008 difference). The standard method produces a lower limit below zero (impossible for proportions), while the arcsine transformation keeps limits in the valid [0,1] range. The arcsine method is more appropriate here.

<br>
<br>

## Case 3: Very Extreme Proportions (Substantial Divergence)

With proportions very close to 0 or 1, the standard method becomes unreliable.

```{r}
#| echo: false
#| warning: false

# Case 3: Very extreme proportions
p1_case3 <- 0.01
p2_case3 <- 0.03
n1_case3 <- 500
n2_case3 <- 500

# Standard method
overall_p_case3 <- (p1_case3 * n1_case3 + p2_case3 * n2_case3) / (n1_case3 + n2_case3)
se_standard_case3 <- sqrt(overall_p_case3 * (1 - overall_p_case3) * (1/n1_case3 + 1/n2_case3))
upper_standard_case3 <- overall_p_case3 + z_crit * se_standard_case3
lower_standard_case3 <- overall_p_case3 - z_crit * se_standard_case3

# Arcsine method
trans1_case3 <- asin(sqrt(p1_case3))
trans2_case3 <- asin(sqrt(p2_case3))
overall_trans_case3 <- asin(sqrt(overall_p_case3))
se_arcsin_case3 <- sqrt(1 / (4 * mean(c(n1_case3,
                                         n2_case3))))
upper_trans_case3 <- overall_trans_case3 + z_crit * se_arcsin_case3
lower_trans_case3 <- overall_trans_case3 - z_crit * se_arcsin_case3
upper_arcsin_case3 <- sin(upper_trans_case3)^2
lower_arcsin_case3 <- sin(lower_trans_case3)^2

# Create comparison data
method_compare_case3 <- data.frame(method = c("Standard",
                                               "Standard",
                                               "Arcsine",
                                               "Arcsine"),
                                   limit_type = c("Upper",
                                                  "Lower",
                                                  "Upper",
                                                  "Lower"),
                                   value = c(upper_standard_case3,
                                            lower_standard_case3,
                                            upper_arcsin_case3,
                                            lower_arcsin_case3),
                                   valid = c("Yes",
                                            "NO - Negative!",
                                            "Yes",
                                            "Yes"))

method_compare_case3 |>
  flextable::flextable() |>
  flextable::set_header_labels(method = "Method",
                                limit_type = "Limit",
                                value = "Value",
                                valid = "Valid Range?") |>
  flextable::colformat_double(j = "value",
                               digits = 4) |>
  flextable::theme_booktabs() |>
  flextable::autofit() |>
  flextable::align(align = "left",
                   part = "all") |>
  flextable::bold(part = "header") |>
  flextable::bg(i = ~ valid == "NO - Negative!",
                bg = "#FFB6C1",
                part = "body") |>
  flextable::add_header_row(values = c("",
                                       "",
                                       "Case 3: Very Extreme Proportions (p ≈ 0.01-0.03)"),
                             colwidths = c(1, 1, 2))

```

<br>

**Observation**: The standard method completely breaks down, producing a negative lower limit (-0.0022), which is impossible for proportions. The arcsine transformation maintains valid bounds and provides accurate inference. **Always use the arcsine transformation for proportions near 0 or 1.**

<br>
<br>

## Visual Comparison: Standard vs. Arcsine Methods

```{r}
#| echo: false
#| warning: false
#| fig-height: 8

# Create data for visualization
prop_values <- seq(0.01,
                   0.99,
                   by = 0.01)
n_sample <- 500

# Calculate SE for both methods
se_standard_vec <- sqrt(prop_values * (1 - prop_values) / n_sample)
se_arcsin_vec <- sqrt(1 / (4 * n_sample))

comparison_plot_df <- data.frame(proportion = rep(prop_values,
                                                   2),
                                 standard_error = c(se_standard_vec,
                                                   rep(se_arcsin_vec[1],
                                                       length(prop_values))),
                                 method = rep(c("Standard",
                                               "Arcsine"),
                                             each = length(prop_values)))

ggplot2::ggplot(data = comparison_plot_df,
                mapping = ggplot2::aes(x = proportion,
                                       y = standard_error,
                                       color = method,
                                       linetype = method)) +
  ggplot2::geom_line(linewidth = 1.5) +
  ggplot2::annotate(geom = "rect",
                    xmin = 0,
                    xmax = 0.15,
                    ymin = 0,
                    ymax = max(comparison_plot_df$standard_error),
                    alpha = 0.2,
                    fill = "red") +
  ggplot2::annotate(geom = "rect",
                    xmin = 0.85,
                    xmax = 1,
                    ymin = 0,
                    ymax = max(comparison_plot_df$standard_error),
                    alpha = 0.2,
                    fill = "red") +
  ggplot2::annotate(geom = "text",
                    x = 0.075,
                    y = 0.020,
                    label = "Use Arcsine",
                    size = 5,
                    angle = 90) +
  ggplot2::annotate(geom = "text",
                    x = 0.925,
                    y = 0.020,
                    label = "Use Arcsine",
                    size = 5,
                    angle = 90) +
  ggplot2::scale_color_manual(values = c("Arcsine" = "blue",
                                         "Standard" = "darkgreen")) +
  ggplot2::labs(title = "When to Use Arcsine Transformation for Proportions",
                subtitle = "Standard Error Comparison (n = 500)",
                x = "Proportion",
                y = "Standard Error",
                color = "Method",
                linetype = "Method") +
  ggplot2::theme(legend.position = "bottom")

```

<br>
<br>

**Key Takeaways for Method Selection**:

1. **Use Standard Method**: When proportions are between 0.15 and 0.85 (shaded regions excluded)

2. **Use Arcsine Method**: When proportions are below 0.15 or above 0.85 (red shaded regions)

3. **Why it matters**: The standard method assumes constant variance across the proportion range, but variance actually depends on p (being highest at p=0.5 and lowest near 0 and 1). The arcsine transformation stabilizes this variance.

4. **Practical recommendation**: When in doubt, use the arcsine transformation—it works well across all proportion ranges and never produces invalid results.











